## Reproducible Research Course Project 2

Data Processing

```{r loadlibraries}
library(ggplot2)  
library(plyr) 
```
Loading the data
The data was downloaded from the link above and saved on local computer (in setwd command one can replace loacal file path with path of folder where the data was downloaded). Then it was loaded on the R using the read.csv command. If object strom.data is already loaded, use that cached object insted of loading it each time the Rmd file is knitted.

```{r load data}
library(readr)
storm.data <- read_csv("/Applications/Documents/repdata-data-StormData.csv.bz2")

```
Examine Structure of data
 
```{r examine}
dim(storm.data)
```
Extracting variables of interest for analysis of weather impact on health and economy
From a list of variables in storm.data, these are columns of interest:

Health variables:

FATALITIES: approx. number of deaths
INJURIES: approx. number of injuries
Economic variables:

PROPDMG: approx. property damags
PROPDMGEXP: the units for property damage value
CROPDMG: approx. crop damages
CROPDMGEXP: the units for crop damage value
Events - target variable:

EVTYPE: weather event (Tornados, Wind, Snow, Flood, etc..)
Extract variables of interest from original data set:

```{r var}
vars <- c( "EVTYPE", "FATALITIES", "INJURIES", "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP")
mydata <- storm.data[, vars]
```
Check the last few rows in data set (in firs years of recording there are many missing (NA) values):
```{r tail}
tail(mydata)
```
Checking for missing values
In every analysis we must the check number of missing values in variables.

Check for missing values in health variables - there is no NA’s in the data.
```{R MissingNA}
sum(is.na(mydata$FATALITIES))
```

```{R MissingNA1}
sum(is.na(mydata$INJURIES))
```

Check for missing values in economic variables for “size” of damage - there is no NA’s in the data.
```{R MissingEconomic}
sum(is.na(mydata$PROPDMG))
```

```{R MissingEconomic1}
sum(is.na(mydata$CROPDMG))
```
Check for missing values in economic variables for units damage - there is no NA’s in the data.
```{R MissingEconomic2}
sum(is.na(mydata$PROPDMGEXP))
```

```{R MissingEconomic3}
sum(is.na(mydata$CROPDMGEXP))
```

Transforming extracted variables
Listing the first 10 event types that most appear in the data:
```{R ExtractedVars}
sort(table(mydata$EVTYPE), decreasing = TRUE)[1:10]
```

We will group events like TUNDERSTORM WIND, TUNDERSTORM WINDS, HIGH WIND, etc. by containing the keyword ‘WIND’ as one event WIND. And we will transform other types of events in a similar way. New variable EVENTS is the transform variable of EVTYPE that have 10 different types of events: HEAT, FLOOD, etc., and type OTHER for events in which name the keyword is not found.

```{R Event}
# create a new variable EVENT to transform variable EVTYPE in groups
mydata$EVENT <- "OTHER"
# group by keyword in EVTYPE
mydata$EVENT[grep("HAIL", mydata$EVTYPE, ignore.case = TRUE)] <- "HAIL"
mydata$EVENT[grep("HEAT", mydata$EVTYPE, ignore.case = TRUE)] <- "HEAT"
mydata$EVENT[grep("FLOOD", mydata$EVTYPE, ignore.case = TRUE)] <- "FLOOD"
mydata$EVENT[grep("WIND", mydata$EVTYPE, ignore.case = TRUE)] <- "WIND"
mydata$EVENT[grep("STORM", mydata$EVTYPE, ignore.case = TRUE)] <- "STORM"
mydata$EVENT[grep("SNOW", mydata$EVTYPE, ignore.case = TRUE)] <- "SNOW"
mydata$EVENT[grep("TORNADO", mydata$EVTYPE, ignore.case = TRUE)] <- "TORNADO"
mydata$EVENT[grep("WINTER", mydata$EVTYPE, ignore.case = TRUE)] <- "WINTER"
mydata$EVENT[grep("RAIN", mydata$EVTYPE, ignore.case = TRUE)] <- "RAIN"
# listing the transformed event types 
sort(table(mydata$EVENT), decreasing = TRUE)
```
Checking the values for variables that represent units od dollars:
```{R Sort}
sort(table(mydata$PROPDMGEXP), decreasing = TRUE)[1:10]
```

```{R Sort1}
sort(table(mydata$CROPDMGEXP), decreasing = TRUE)[1:10]
```
There is some mess in units, so we transform those variables in one unit (dollar) variable by the following rule:

K or k: thousand dollars (10^3)

M or m: million dollars (10^6)

B or b: billion dollars (10^9)

the rest would be consider as dollars

New variable(s) is product of value of damage and dollar unit.
```{R Event2}
mydata$PROPDMGEXP <- as.character(mydata$PROPDMGEXP)
mydata$PROPDMGEXP[is.na(mydata$PROPDMGEXP)] <- 0 # NA's considered as dollars
mydata$PROPDMGEXP[!grepl("K|M|B", mydata$PROPDMGEXP, ignore.case = TRUE)] <- 0 # everything exept K,M,B is dollar
mydata$PROPDMGEXP[grep("K", mydata$PROPDMGEXP, ignore.case = TRUE)] <- "3"
mydata$PROPDMGEXP[grep("M", mydata$PROPDMGEXP, ignore.case = TRUE)] <- "6"
mydata$PROPDMGEXP[grep("B", mydata$PROPDMGEXP, ignore.case = TRUE)] <- "9"
mydata$PROPDMGEXP <- as.numeric(as.character(mydata$PROPDMGEXP))
mydata$property.damage <- mydata$PROPDMG * 10^mydata$PROPDMGEXP

mydata$CROPDMGEXP <- as.character(mydata$CROPDMGEXP)
mydata$CROPDMGEXP[is.na(mydata$CROPDMGEXP)] <- 0 # NA's considered as dollars
mydata$CROPDMGEXP[!grepl("K|M|B", mydata$CROPDMGEXP, ignore.case = TRUE)] <- 0 # everything exept K,M,B is dollar
mydata$CROPDMGEXP[grep("K", mydata$CROPDMGEXP, ignore.case = TRUE)] <- "3"
mydata$CROPDMGEXP[grep("M", mydata$CROPDMGEXP, ignore.case = TRUE)] <- "6"
mydata$CROPDMGEXP[grep("B", mydata$CROPDMGEXP, ignore.case = TRUE)] <- "9"
mydata$CROPDMGEXP <- as.numeric(as.character(mydata$CROPDMGEXP))
mydata$crop.damage <- mydata$CROPDMG * 10^mydata$CROPDMGEXP
```
Print of first 10 values for property damage (in dollars) that most appear in the data:
```{R sort2}
sort(table(mydata$property.damage), decreasing = TRUE)[1:10]
```
Print of first 10 values for crop damage (in dollars) that most appear in the data:

```{R cropdamage}
sort(table(mydata$property.damage), decreasing = TRUE)[1:10]
```

Analysis
Aggregating events for public health variables
Table of public health problems by event type
```{R Analysis}
# aggregate FATALITIES and INJURIES by type of EVENT
agg.fatalites.and.injuries <- ddply(mydata, .(EVENT), summarize, Total = sum(FATALITIES + INJURIES,  na.rm = TRUE))
agg.fatalites.and.injuries$type <- "fatalities and injuries"
  
# aggregate FATALITIES by type of EVENT
agg.fatalities <- ddply(mydata, .(EVENT), summarize, Total = sum(FATALITIES, na.rm = TRUE))
agg.fatalities$type <- "fatalities"

# aggregate INJURIES by type of EVENT
agg.injuries <- ddply(mydata, .(EVENT), summarize, Total = sum(INJURIES, na.rm = TRUE))
agg.injuries$type <- "injuries"

# combine all
agg.health <- rbind(agg.fatalities, agg.injuries)

health.by.event <- join (agg.fatalities, agg.injuries, by="EVENT", type="inner")
health.by.event
```

Aggregating events for economic variables

```{R Aggreg}
# aggregate PropDamage and CropDamage by type of EVENT
agg.propdmg.and.cropdmg <- ddply(mydata, .(EVENT), summarize, Total = sum(property.damage + crop.damage,  na.rm = TRUE))
agg.propdmg.and.cropdmg$type <- "property and crop damage"

# aggregate PropDamage by type of EVENT
agg.prop <- ddply(mydata, .(EVENT), summarize, Total = sum(property.damage, na.rm = TRUE))
agg.prop$type <- "property"

# aggregate INJURIES by type of EVENT
agg.crop <- ddply(mydata, .(EVENT), summarize, Total = sum(crop.damage, na.rm = TRUE))
agg.crop$type <- "crop"

# combine all
agg.economic <- rbind(agg.prop, agg.crop)


economic.by.event <- join (agg.prop, agg.crop, by="EVENT", type="inner")
economic.by.event
```

Results
Across the United States, which types of events are most harmful with respect to population health?
```{R Results}
# transform EVENT to factor variable for health variables
agg.health$EVENT <- as.factor(agg.health$EVENT)

# plot FATALITIES and INJURIES by EVENT
health.plot <- ggplot(agg.health, aes(x = EVENT, y = Total, fill = type)) + geom_bar(stat = "identity") +
  coord_flip() +
  xlab("Event Type") + 
  ylab("Total number of health impact") +
  ggtitle("Weather event types impact on public health") +
  theme(plot.title = element_text(hjust = 0.5))
print(health.plot)  
```

Across the United States, which types of events have the greatest economic consequences?
```{R economic}
# # transform EVENT to factor variable for economic variables
agg.economic$EVENT <- as.factor(agg.economic$EVENT)

# plot PROPERTY damage and CROP damage by EVENT
economic.plot <- ggplot(agg.economic, aes(x = EVENT, y = Total, fill = type)) + geom_bar(stat = "identity") +
  coord_flip() +
  xlab("Event Type") + 
  ylab("Total damage in dollars") +
  ggtitle("Weather event types impact on property and crop damage") +
  theme(plot.title = element_text(hjust = 0.5))
print(economic.plot) 
```

